The Unicode Cookbook \linebreak for Linguists

### 2018-03-06 book/unicode.tex
these changes almost complete the remaining reviews

@@

\include{chapters/introduction}
\include{chapters/pitfalls}
\include{chapters/ipa_background}
\include{chapters/ipa_meets_unicode}
\include{chapters/practical_recommendations}
\include{chapters/orthography_profiles}

@@

To make the functionality even more accessible, we have prepared webapps with 
the \texttt{shiny} framework for the R functions. These webapps are available 
online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
package and can be started with the following helper function:

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{launch_shiny}\hlstd{(}\hlstr{'tokenize'}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

% TO DO

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Feb  3 16:28:50 2018
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ú & 1 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Ù & 1 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
  Û & 1 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 1 (default settings, splitting grapheme clusters)} 
\label{tab:profile1}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Feb  3 16:28:50 2018
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 1 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 1 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 1 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 3 & U+0055 & LATIN CAPITAL LETTER U \\ 
  Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 2 (splitting by codepoints)} 
\label{tab:profile2}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Feb  3 16:28:51 2018
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 3 (splitting by NFC codepoints)} 
\label{tab:profile3}
\end{table}

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Feb  3 16:28:51 2018
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 2 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 2 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 2 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 6 & U+0055 & LATIN CAPITAL LETTER U \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 4 (splitting by NFD codepoints)} 
\label{tab:profile4}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Feb  3 16:28:51 2018
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 2 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 2 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 5 (splitting by graphemes after NFD)} 
\label{tab:profile5}
\end{table}

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Feb  3 16:28:51 2018
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 6 (splitting by graphemes after NFC)} 
\label{tab:profile6}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Feb  3 16:28:51 2018
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & UnicodeName \\ 
  \midrule
sh &  &  &  \\ 
  ch &  &  &  \\ 
  sch &  &  &  \\ 
  ie &  &  &  \\ 
  oo &  &  &  \\ 
  a & 2 & U+0061 & LATIN SMALL LETTER A \\ 
  c & 2 & U+0063 & LATIN SMALL LETTER C \\ 
  e & 3 & U+0065 & LATIN SMALL LETTER E \\ 
  f & 1 & U+0066 & LATIN SMALL LETTER F \\ 
  h & 4 & U+0068 & LATIN SMALL LETTER H \\ 
  i & 3 & U+0069 & LATIN SMALL LETTER I \\ 
  m & 4 & U+006D & LATIN SMALL LETTER M \\ 
  o & 2 & U+006F & LATIN SMALL LETTER O \\ 
  p & 2 & U+0070 & LATIN SMALL LETTER P \\ 
  s & 4 & U+0073 & LATIN SMALL LETTER S \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Manually edited profile skeleton} 
\label{tab:profile_skeleton1}
\end{table}

It is also possible to leave out the Unicode information in
the profile skeleton by using the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simple situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Feb  3 16:28:51 2018
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{lllll}
  \toprule
Left & Grapheme & Frequency & Codepoint & UnicodeName \\ 
  \midrule
mi & s &  &  &  \\ 
   & sh &  &  &  \\ 
   & ch &  &  &  \\ 
   & sch &  &  &  \\ 
   & ie &  &  &  \\ 
   & oo &  &  &  \\ 
   & a & 2 & U+0061 & LATIN SMALL LETTER A \\ 
   & c & 2 & U+0063 & LATIN SMALL LETTER C \\ 
   & e & 3 & U+0065 & LATIN SMALL LETTER E \\ 
   & f & 1 & U+0066 & LATIN SMALL LETTER F \\ 
   & h & 4 & U+0068 & LATIN SMALL LETTER H \\ 
   & i & 3 & U+0069 & LATIN SMALL LETTER I \\ 
   & m & 4 & U+006D & LATIN SMALL LETTER M \\ 
   & o & 2 & U+006F & LATIN SMALL LETTER O \\ 
   & p & 2 & U+0070 & LATIN SMALL LETTER P \\ 
   & s & 4 & U+0073 & LATIN SMALL LETTER S \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Orthography profile with contextual specification for <s>} 
\label{tab:profile_skeleton2}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Feb  3 16:28:51 2018
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{lllll}
  \toprule
Left & Grapheme & Right & Class & Replacement \\ 
  \midrule
 & a &  &  & a \\ 
   & c &  &  & c \\ 
   & e &  &  & e \\ 
   & i &  &  & i \\ 
   & n &  &  & n \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Orthography profile with empty columns for editing contexts} 
\label{tab:profile_editing_1}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Feb  3 16:28:51 2018
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{lllll}
  \toprule
Left & Grapheme & Right & Class & IPA \\ 
  \midrule
 & c & [ie] &  & tʃ \\ 
   & a &  &  & a \\ 
   & n &  &  & n \\ 
   & c &  &  & k \\ 
   & e &  &  & e \\ 
   & i &  &  & i \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Orthography profile with regex as context} 
\label{tab:profile_editing_2}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Feb  3 16:28:51 2018
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{lllll}
  \toprule
Left & Grapheme & Right & Class & IPA \\ 
  \midrule
 & c & Vfront &  & tʃ \\ 
   & a &  &  & a \\ 
   & n &  &  & n \\ 
   & c &  &  & k \\ 
   & e &  & Vfront & e \\ 
   & i &  & Vfront & i \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Orthography profile with Class as context} 
\label{tab:profile_editing_3}
\end{table}

>>>>

\lstset{
  literate={ĉ}{{\^c}}1
           {á}{{\'a}}1
           {ã̌}{{\'a}}1
}

@@

\include{chapters/introduction}
\include{chapters/unicode_approach}
\include{chapters/pitfalls}
\include{chapters/ipa_background}
\include{chapters/ipa_meets_unicode}
\include{chapters/practical_recommendations}
\include{chapters/orthography_profiles}

@@

% TODO
% To make the functionality even more accessible, we have prepared webapps with 
% the \texttt{shiny} framework for the R functions. These webapps are available 
% online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
% package and can be started with the following helper function:

% <<eval=FALSE>>=
% launch_shiny('tokenize')
% @

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Tue Mar  6 15:25:08 2018
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ú & 1 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Ù & 1 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
  Û & 1 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 1 (default settings, splitting grapheme clusters)} 
\label{tab:profile1}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Tue Mar  6 15:25:08 2018
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 1 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 1 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 1 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 3 & U+0055 & LATIN CAPITAL LETTER U \\ 
  Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 2 (splitting by codepoints)} 
\label{tab:profile2}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Tue Mar  6 15:25:08 2018
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 3 (splitting by NFC codepoints)} 
\label{tab:profile3}
\end{table}

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Tue Mar  6 15:25:08 2018
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 2 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 2 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 2 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 6 & U+0055 & LATIN CAPITAL LETTER U \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 4 (splitting by NFD codepoints)} 
\label{tab:profile4}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Tue Mar  6 15:25:08 2018
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 2 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 2 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 5 (splitting by graphemes after NFD)} 
\label{tab:profile5}
\end{table}

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Tue Mar  6 15:25:08 2018
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 6 (splitting by graphemes after NFC)} 
\label{tab:profile6}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Tue Mar  6 15:25:08 2018
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & UnicodeName \\ 
  \midrule
sh &  &  &  \\ 
  ch &  &  &  \\ 
  sch &  &  &  \\ 
  ie &  &  &  \\ 
  oo &  &  &  \\ 
  a & 2 & U+0061 & LATIN SMALL LETTER A \\ 
  c & 2 & U+0063 & LATIN SMALL LETTER C \\ 
  e & 3 & U+0065 & LATIN SMALL LETTER E \\ 
  f & 1 & U+0066 & LATIN SMALL LETTER F \\ 
  h & 4 & U+0068 & LATIN SMALL LETTER H \\ 
  i & 3 & U+0069 & LATIN SMALL LETTER I \\ 
  m & 4 & U+006D & LATIN SMALL LETTER M \\ 
  o & 2 & U+006F & LATIN SMALL LETTER O \\ 
  p & 2 & U+0070 & LATIN SMALL LETTER P \\ 
  s & 4 & U+0073 & LATIN SMALL LETTER S \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Manually edited profile skeleton} 
\label{tab:profile_skeleton1}
\end{table}

To leave out the Unicode information in
the profile skeleton, use the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simple situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Tue Mar  6 15:25:09 2018
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{lllll}
  \toprule
Left & Grapheme & Frequency & Codepoint & UnicodeName \\ 
  \midrule
mi & s &  &  &  \\ 
   & sh &  &  &  \\ 
   & ch &  &  &  \\ 
   & sch &  &  &  \\ 
   & ie &  &  &  \\ 
   & oo &  &  &  \\ 
   & a & 2 & U+0061 & LATIN SMALL LETTER A \\ 
   & c & 2 & U+0063 & LATIN SMALL LETTER C \\ 
   & e & 3 & U+0065 & LATIN SMALL LETTER E \\ 
   & f & 1 & U+0066 & LATIN SMALL LETTER F \\ 
   & h & 4 & U+0068 & LATIN SMALL LETTER H \\ 
   & i & 3 & U+0069 & LATIN SMALL LETTER I \\ 
   & m & 4 & U+006D & LATIN SMALL LETTER M \\ 
   & o & 2 & U+006F & LATIN SMALL LETTER O \\ 
   & p & 2 & U+0070 & LATIN SMALL LETTER P \\ 
   & s & 4 & U+0073 & LATIN SMALL LETTER S \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Orthography profile with contextual specification for <s>} 
\label{tab:profile_skeleton2}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Tue Mar  6 15:25:09 2018
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{lllll}
  \toprule
Left & Grapheme & Right & Class & Replacement \\ 
  \midrule
 & a &  &  & a \\ 
   & c &  &  & c \\ 
   & e &  &  & e \\ 
   & i &  &  & i \\ 
   & n &  &  & n \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Orthography profile with empty columns for editing contexts} 
\label{tab:profile_editing_1}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Tue Mar  6 15:25:09 2018
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{lllll}
  \toprule
Left & Grapheme & Right & Class & IPA \\ 
  \midrule
 & c & [ie] &  & tʃ \\ 
   & a &  &  & a \\ 
   & n &  &  & n \\ 
   & c &  &  & k \\ 
   & e &  &  & e \\ 
   & i &  &  & i \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Orthography profile with regex as context} 
\label{tab:profile_editing_2}
\end{table}

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Tue Mar  6 15:25:09 2018
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{lllll}
  \toprule
Left & Grapheme & Right & Class & IPA \\ 
  \midrule
 & c & Vfront &  & tʃ \\ 
   & a &  &  & a \\ 
   & n &  &  & n \\ 
   & c &  &  & k \\ 
   & e &  & Vfront & e \\ 
   & i &  & Vfront & i \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Orthography profile with Class as context} 
\label{tab:profile_editing_3}
\end{table}
